---
title: "Weather and Traffic Accidents in Toronto"
author: "Kaiwen Zhang"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE, error = FALSE)
library(tidyverse)
library(ggplot2)
library(kableExtra)
library(lubridate)
library(dplyr)
```

# Project Overview

This project investigates how weather conditions—specifically temperature and precipitation—impact the frequency and severity of traffic accidents in Toronto between 2006 and 2023. Using data from the Toronto Police Service and ERA5 reanalysis archive, we merged accident records with historical weather data to assess correlations and predictive potential.

[📄 **Download Full Report (PDF)**](Final.pdf)

---

## Table – Summary of Dataset
```{r table_dataset_summary}
df_final <- read.csv("data/df_final.csv")
tibble(
  "Total Observations" = nrow(df_final),
  "Date Range" = paste(min(df_final$DATE), "to", max(df_final$DATE)),
  "Variables" = ncol(df_final)
) %>%
  kable("html", caption = "Table 1: Dataset Overview") %>%
  kable_styling(full_width = FALSE)
```

---

# Key Findings

- **Seasonality:** Accidents peak in warmer months, especially August.
- **Temperature:** A statistically significant but modest positive relationship exists between temperature and accident frequency.
- **Precipitation:** Rainy days see fewer accidents, but injury severity is similar to dry days.
- **Prediction:** Random Forest models using only weather and surface conditions yield limited predictive power for injury severity.

---

# Highlighted Visualizations

## Fig 1 – Monthly Accident Totals
```{r fig1_monthly_accidents}
df_final %>%
  mutate(MONTH = lubridate::month(DATE, label = TRUE, locale = "en_US")) %>% 
  group_by(MONTH) %>%
  summarise(Total_Accidents = n()) %>%
  ggplot(aes(x = MONTH, y = Total_Accidents, fill = MONTH)) +
  geom_bar(stat = "identity") +
  labs(title = "Monthly Accident Totals", x = "Month", y = "Number of Accidents") +
  theme_minimal()
```

Accidents increase in summer months, peaking in August. This may reflect higher road usage.

---

## Fig 2 – Temperature vs Fatalities
```{r fig2_temp_fatal}
df_final %>%
  filter(FATAL_NO > 0) %>%
  mutate(TEMP_AVG = (TEMP_MIN + TEMP_MAX) / 2) %>%
  mutate(FATAL_TYPE = ifelse(FATAL_NO >= 30, "High Fatal", "Low Fatal")) %>%
  ggplot(aes(x = TEMP_AVG, y = FATAL_NO, color = FATAL_TYPE)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Fig 2b - Stratified Fatal Accidents by Temperature",
       x = "Average Temperature (°C)",
       y = "Number of Fatalities") +
  theme_minimal()
```

There is a mild positive association between temperature and the number of fatalities.

---

## Fig 3 – Injury Severity by Rain
```{r fig3_injury_rain}
ggplot(df_final, aes(x = RAINY_DAY, fill = INJURY)) +
  geom_bar(position = "stack") +
  labs(title = "Injury Levels on Rainy vs Dry Days",
       x = "Weather Condition", y = "Accident Count") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set2")
```

Rainy days have fewer total accidents but similar injury severity distribution as dry days.

---

## Fig 4 – Variable Importance (Random Forest)
```{r fig4_varimp, message=FALSE}
library(randomForest)
rf_model <- randomForest(INJURY_BIN ~ TEMP_MIN + TEMP_MAX + PRECIPITATION + VISIBILITY + LIGHT + RDSFCOND + RAINY_DAY,
                         data = df_final %>% filter(INJURY != "Unknown") %>% mutate(INJURY_BIN = as.factor(ifelse(INJURY %in% c("Fatal", "Major"), "Severe", "NotSevere"))),
                         ntree = 100, importance = TRUE)
varImpPlot(rf_model, main = "Variable Importance from Random Forest")
```

This suggests that while temperature-related variables are important structurally in the model, rain indicators and visibility are more directly tied to predictive accuracy.

---

# Conclusion

While weather conditions do show some correlation with traffic accident frequency and severity, the predictive power of weather-related variables alone is limited. Future work should consider integrating traffic volume, driver behavior, and spatial patterns to improve predictive accuracy.

Visit the [interactive visualizations page](visualizations.html) for dynamic maps and filterable charts.
